Halving the cost of quantum addition

### 2017-10-04 paper.tex
Generating figures as separate PDFs

\input{Qcircuit}
\newcommand{\qH}{\gate{H}}
\newcommand{\qT}{\gate{T}}
\newcommand{\qTi}{\gate{T^\dagger}}
\newcommand{\qS}{\gate{S}}
\newcommand{\qSi}{\gate{S^\dagger}}
\newcommand{\qA}{\lstick{|A\rangle}}
\newcommand{\qO}{\lstick{|0\rangle}}

@@

\begin{figure}
  \input{figures/adder.qcircuit}
  \caption{
	A 5-bit adder with T-count of 16.
	Uses Clifford operations, four logical-AND computations each with a T-count of 4, and four logical-AND erasures requiring no T gates.
	Generalizes to an $n$-bit adder with a T-count of $4n - 4$.
	See \autoref{fig:full-adder-block} for the adder building-block and \autoref{fig:temporary-logical-AND} for the logical-AND computation and uncomputation circuits.
  }
  \label{fig:multi-bit-adder-example}
\end{figure}

@@

In \autoref{fig:multi-bit-adder-example}, we present a 5-bit adder with a T-count of 16.
It performs 4 temporary logical-ANDs, each with a T-count of 4.
All other operations are Clifford operations, with no T-count.

The building block of our adder is shown in \autoref{fig:full-adder-block}.
We construct $n$-bit adders by nesting $n$ copies of the building block inside of each other.
The outer-most and inner-most blocks (which act on the low bit and high bit respectively) are then specialized based on the fact that they either have no carry input or no carry output.

\begin{figure}
  \input{figures/adder-building-block.qcircuit}
  \caption{
	Our adder circuit building-block, with a T-count of 4.
	A variant of the Cuccaro adder building-block \citep{Cuccaro2004}.
  }
  \label{fig:full-adder-block}
\end{figure}

@@

\autoref{fig:variations-on-full-adder-block} shows the building-blocks for two variations on our adder: a controlled adder and a temporary adder.

@@

\begin{figure}
  \input{figures/temporary-logical-and.qcircuit}
  \caption{
	How to compute and uncompute the logical-AND of two qubits.
	The computation circuit (top) has a T-count of 4 and a T-depth of 1.
	Note that the $|A\rangle$ state input contributes to the T-count, because $|A\rangle$ states are the resource used to perform T gates.
	The uncomputation circuit (bottom) uses a measure-and-fixup approach \cite{Jones2013} that requires only Clifford gates, and so has a T-count of zero.
    \\
    An alternative uncomputation construction is to simply do the reverse of the computation circuit.
    This alternative approach has a net T-count of 2 (because an $|A\rangle$ state is recovered).
    The resulting temporary logical-AND with a T-count of 6 would still be an improvement on existing work, but would be inferior to the measure-and-fixup approach shown above.
  }
  \label{fig:temporary-logical-AND}
\end{figure}

@@

\begin{figure}
  \input{figures/other-adder-building-blocks.qcircuit}
  \caption{
	Variations on our adder construction.
	(a) is a controlled-adder building-block with a T-count of 8.
	The addition will only occur when the control is on.
	(b) is a temporary-adder building-block with a T-count of 4.
	Has the same T-count as a normal addition, but makes the result available before any uncomputation steps are needed.
  }
  \label{fig:variations-on-full-adder-block}
\end{figure}

The temporary logical-AND is useful for optimizing an even wider variety of circuits than addition is.
Whenever Toffoli gates appear in compute/uncompute pairs, and intermediate operations are not sensitive to phase errors on the controls of the Toffoli gate (i.e. the condition in \autoref{fig:pair-definition} is satisfied), it is possible to save 4 T gates by replacing the pair of Toffoli gates with a temporary logical-AND.

\begin{figure}
  \input{figures/paired-toffoli-replacement-rule.qcircuit}
  \caption{
	A sufficient condition for replacing a pair of Toffoli gates with a temporary logical-AND, saving 4 T gates.\\
	1) The later Toffoli gate must be uncomputing the earlier Toffoli gate.\\
	2) Intermediate operations must not be sensitive to the presence of the entangled ancilla.
  }
  \label{fig:pair-definition}
\end{figure}

@@

It is interesting to consider if addition or temporary logical-ANDs can be done with still fewer T gates.
\autoref{fig:lower-bound-logical-AND} presents a simple proof that the T-count of the temporary logical-AND is at least 2.
Although we suspect the true lower bound is 4, we also expect that there are special cases with lower T-counts.
For example, if a task involves repeating an action several times (e.g. counting by repeatedly conditionally incrementing a register), T gates from Toffolis in uncomputation work at the end of one repetition may cancel against T gates from Toffolis in computation work at the start of the next repetition.

\begin{figure}
  \input{figures/synthesize-t-gates-from-temporary-logical-and.qcircuit}
  \caption{
	Three T gates can be synthesized by performing a temporary logical-AND, one T gate, and various Clifford operations.
	Clifford operations can't produce T gates, so the extra T gates must be coming from the temporary logical-AND.
	This proves that performing the temporary logical-AND requires at least two T gates.
  }
  \label{fig:lower-bound-logical-AND}
\end{figure}

@@

\section*{Methods}

The described circuits can be verified with a quantum circuit simulator, such as ProjectQ or LIQUID or Quirk.

@@

\section*{Competing Interests}

No competing interests.

\section*{Contributions}

Craig Gidney discovered and refined the temporary logical-AND construction, applied it to adders and the other discussed circuits, and is the sole author of the paper.

\section*{Funding}

Work was done while employed at Google.

>>>>

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/adder.pdf}
  \caption{
	A 5-bit adder with T-count of 16.
	Uses Clifford operations, four logical-AND computations each with a T-count of 4, and four logical-AND erasures requiring no T gates.
	Generalizes to an $n$-bit adder with a T-count of $4n - 4$.
	See \autoref{fig:adder-building-block} for the adder building-block and \autoref{fig:temporary-logical-AND} for the logical-AND computation and uncomputation circuits.
  }
  \label{fig:adder}
\end{figure}

@@

In \autoref{fig:adder}, we present a 5-bit adder with a T-count of 16.
It performs 4 temporary logical-ANDs, each with a T-count of 4.
All other operations are Clifford operations, with no T-count.

The building block of our adder is shown in \autoref{fig:adder-building-block}.
We construct $n$-bit adders by nesting $n$ copies of the building block inside of each other.
The outer-most and inner-most blocks (which act on the low bit and high bit respectively) are then specialized based on the fact that they either have no carry input or no carry output.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/adder-building-block.pdf}
  \caption{
	Our adder circuit building-block, with a T-count of 4.
	A variant of the Cuccaro adder building-block \citep{Cuccaro2004}.
  }
  \label{fig:adder-building-block}
\end{figure}

@@

\autoref{fig:other-adder-building-blocks} shows the building-blocks for two variations on our adder: a controlled adder and a temporary adder.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/temporary-logical-and.pdf}
  \caption{
	How to compute and uncompute the logical-AND of two qubits.
	The computation circuit (top) has a T-count of 4 and a T-depth of 1.
	Note that the $|A\rangle$ state input contributes to the T-count, because $|A\rangle$ states are the resource used to perform T gates.
	The uncomputation circuit (bottom) uses a measure-and-fixup approach \cite{Jones2013} that requires only Clifford gates, and so has a T-count of zero.
    \\
    An alternative uncomputation construction is to simply do the reverse of the computation circuit.
    This alternative approach has a net T-count of 2 (because an $|A\rangle$ state is recovered).
    The resulting temporary logical-AND with a T-count of 6 would still be an improvement on existing work, but would be inferior to the measure-and-fixup approach shown above.
  }
  \label{fig:temporary-logical-AND}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/other-adder-building-blocks.pdf}
  \caption{
	Variations on our adder construction.
	(a) is a controlled-adder building-block with a T-count of 8.
	The addition will only occur when the control is on.
	(b) is a temporary-adder building-block with a T-count of 4.
	Has the same T-count as a normal addition, but makes the result available before any uncomputation steps are needed.
  }
  \label{fig:other-adder-building-blocks}
\end{figure}

The temporary logical-AND is useful for optimizing an even wider variety of circuits than addition is.
Whenever Toffoli gates appear in compute/uncompute pairs, and intermediate operations are not sensitive to phase errors on the controls of the Toffoli gate (i.e. the condition in \autoref{fig:paired-toffoli-replacement-rule} is satisfied), it is possible to save 4 T gates by replacing the pair of Toffoli gates with a temporary logical-AND.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/paired-toffoli-replacement-rule.pdf}
  \caption{
	A sufficient condition for replacing a pair of Toffoli gates with a temporary logical-AND, saving 4 T gates.\\
	1) The later Toffoli gate must be uncomputing the earlier Toffoli gate.\\
	2) Intermediate operations must not be sensitive to the presence of the entangled ancilla.
  }
  \label{fig:paired-toffoli-replacement-rule}
\end{figure}

@@

It is interesting to consider if addition or temporary logical-ANDs can be done with still fewer T gates.
\autoref{fig:synthesize-t-gates-from-temporary-logical-and} presents a simple proof that the T-count of the temporary logical-AND is at least 2.
Although we suspect the true lower bound is 4, we also expect that there are special cases with lower T-counts.
For example, if a task involves repeating an action several times (e.g. counting by repeatedly conditionally incrementing a register), T gates from Toffolis in uncomputation work at the end of one repetition may cancel against T gates from Toffolis in computation work at the start of the next repetition.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/synthesize-t-gates-from-temporary-logical-and.pdf}
  \caption{
	Three T gates can be synthesized by performing a temporary logical-AND, one T gate, and various Clifford operations.
	Clifford operations can't produce T gates, so the extra T gates must be coming from the temporary logical-AND.
	This proves that performing the temporary logical-AND requires at least two T gates.
  }
  \label{fig:synthesize-t-gates-from-temporary-logical-and}
\end{figure}

@@

@@
